Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-10-23 docs/manuscript/internal_comments_4/difference_file/new.tex
returned package to AM

\usepackage[section]{placeins} %ensures figures go in their section e.g https://tex.stackexchange.com/questions/279/how-do-i-ensure-that-figures-appear-in-the-section-theyre-associated-with

@@

Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$. Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90 \% credible interval of the $3N$ one-dimensional, marginalized posteriors. We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. The subtitles of the marginalized 1D posteriors denote the posterior median and the 0.16,0.84 quantiles. We are able to accurately estimate each parameter of interest. We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. Note that the $x$-axis scaling does not span the total prior space.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes. The subtitles of the marginalized 1D posteriors refer to values for the median posterior across the noise realisations. The injection value is generally contained within the 90 \% credible interval for the vast of majority posteriors. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected Values & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$  sets the bound on the difference in expectation value between the one dimensional posteriors, and is generally small with respect to the width of the prior space.}
	\label{tab:Wasserstein}
\end{table}
\begin{figure*}
	\setkeys{Gin}{width=\linewidth}

	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_0}
		\caption{$\Omega$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_1}
		\caption{$\Phi_0$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_2}
		\caption{$\psi$} \label{figWD_psi}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_3}
		\caption{$\iota$}
	\end{subfigure}
	\medskip
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_4}
		\caption{$\delta$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_5}
		\caption{$\alpha$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_6}
		\caption{$h$}
	\end{subfigure}
	\caption{The first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations, for each parameter of $\boldsymbol{\theta}_{\rm gw}$  $W_1$ provides an upper bound on the difference in expected values between any two probability distributions. $W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$. The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96. An example noise realisation with particularly large pairwise $W_1$ values is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
\end{figure*}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.  The red point labels the location of the maxima, and the magenta point labels the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data. The likelihood has been normalised with respect to the absolute value of the maxima. A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima. This can be solved generally by increasing the number of live points $n_{\rm live}$.}
	\label{fig:likelihood_surface_alpha_psi}
\end{figure}
The results in Section \ref{sec:parameter_estim} are for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$. It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident. It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown. \newline

@@

Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim 10 \%$ (coefficient of variation) across the seven parameters and nine realizations.

 Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval. As in the single noise case, there are tentative signs of bias for some of the parameters, with the one-dimensional posteriors not symmetric about the injected value. For example, the maximum a posteriori probability estimates of $\iota$ across the noise realisations all appear left-skewed, consistently underestimating the injected value. Given the uncertainty in the posteriors it is difficult to draw strong conclusions on a potential bias and this question will be discussed in detail in Section \ref{sec:bias} \newline

We now consider the complete set of $10^3$ noise realisations, rather than the subset of nine realisations in the preceding discussion. For each realisation of the noise, for each parameter, we infer a marginalised one-dimensional posterior. Our goal is to compare these posteriors between different noise realisations and evaluate whether our method can consistently and reliably converge to comparable posteriors. To this end we use the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions. The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}. A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}. The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. Throughout this work we use the first moment of the Wasserstein distance, $W_1(\mu,\nu)$, which is defined between one-dimensional probability distributions $\mu$ and $\nu$ as
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y) \ ,
\end{eqnarray}
where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function. The first moment of the WD has a convenient physical interpretation given by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. We can see that $W_1$ bounds the difference in the expected values. Taking a concrete example, suppose that we infer two one-dimensional posterior distributions for $\iota$, for two different realisations of the noise, and calculate $W_1 =0.5$ rad between them. This means that the difference in the expected value of two random variables drawn from the two posteriors is no greater than $0.5$ rad. \newline

\subsection{Detection} \label{sec:detection}
We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}. $\mathcal{M}_1$ is the Earth-terms only model, i.e. the state-space model with a Kalman filter using Equation \eqref{eq:measuremen_earth}. The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$. The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$. We can see that for $\beta >10$ there is an approximate log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$. We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW. For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3.9 \times 10^{-15}$. This minimal detectable strain is of course particular to this system and will be influenced in general by e.g. $T_{\rm obs}$, $\sigma_{\rm m}$ as well as the parameters of the GW source such as its location on the sky. \newline

There is an evident sparsity and steep drop-off of $\beta$ for values of $h_0 \lesssim 3.9 \times10^{-15}$. The steep drop off is due to the two competing models becoming increasingly indistinguishable as the strain signal decreases and the measurement noise starts to dominate. The sparsity is due some of the $\beta$ value becoming negative at these points.This is a noise artefact of the nested sampling method. For these points the noise in the sampler starts to dominate, the sampler converges sub-optimally, and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails. That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds. There is nothing special about these particular points; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these sparse locations appear at different values of $h_0$, for $\beta < 10$. For $\beta > 10$ the sampling noise is sufficiently small with respect to the signal and the model hierarchy holds. Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$ since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$. For the results presented in this subsection we set $n_{\rm live} =2000$. 
\begin{figure}
	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}. The horizontal grey dashed line labels the detection threshold of $\beta = 10$. The minimal detectable strain, below which $\beta < 10$, is $\sim 3.9 \times 10^{-15}$. Gaps in the data below $\beta=10$ occur due to noise in the sampler breaking the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$.}
	\label{fig:bayes}
\end{figure}

@@

\section{Identifiability and bias} \label{sec:bias_and_identifiability}
There are two underlying issues in the preceding tests that we now explore in more detail. The first relates to parameter identifiability, i.e. can we uniquely identify the parameters of the model given the data and the model structure? This question is explored in Section \ref{sec:identif}. The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}. This is the focus of Section \ref{sec:bias}. In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f. Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}). \footnote{\tiny \textcolor{red}{TK: This discussion on bias and identifiability is perhaps less relevant than it was when our representative example had a 'high' strain ($10^{-12}$). However it IS still a feature/bug of our approach and I think it is worth discussing in the main body rather than relegating to an appendix. Opinions welcome...} \normalsize}

@@

\section{Wasserstein distance}\label{sec:wasserstein}
The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$. It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$. For this reason it is sometimes known as the "Earth mover's distance". The $p$-th order Wasserstein distance between two distributions is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
\end{eqnarray}
where $p \in [1,\infty)$ and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e. $\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$. \newline

@@

The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages. Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another. It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g. radians then we can interpret this as the cost to transform between the two distributions is 1 radian. Further interpretability is provided by the Monge-Kantarovih duality, Equation \eqref{eq:WDdefn}; the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian. Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g. Gaussianity. Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces. Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality. It also respects the geometry of the underlying metric space over which the distributions are defined.

>>>>

%
%\usepackage[section]{placeins} %ensures figures go in their section e.g https://tex.stackexchange.com/questions/279/how-do-i-ensure-that-figures-appear-in-the-section-theyre-associated-with

@@

Similar results are obtained for the 3$N$ parameters in $\boldsymbol{\theta}_{\rm psr, reduced}$. Again, the parameters are recovered unambiguously, in the sense that the nested sampler converges smoothly to a unimodal joint posterior near the known, injected parameter values, with all $3N$ injected parameters lying within the 90\% credible interval of the $3N$ one-dimensional, marginalized posteriors. We do not display the resulting corner plot because it is too big ($3N = 141$), and because inferring ${\boldsymbol{\theta}}_{\rm gw}$ is the focus of this paper and contemporary PTA analyses more generally.

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles. We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
	\caption{As Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors is appreciable, with coefficient of variation $\sim10 \%$ across the 63 combinations. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Weak correlations can be seen between $\Omega - \Phi_0$, $\psi-\alpha$ and $\iota - h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected Values & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & - & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$ is generally smaller than the prior domain.}
	\label{tab:Wasserstein}
\end{table}

The results in Section \ref{sec:parameter_estim} are for a single realisation of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$. It is important to confirm that the analysis scheme returns accurate answers for arbitrary noise realisations and that the specific realisation of the noisy data used in Section \ref{sec:parameter_estim} is not particularly advantageous by accident. It is also important to quantify, albeit approximately, the natural random dispersion in the one-dimensional posterior medians from one noise realization to the next, as the dispersion is a practical measure of the accuracy of the parameter estimation scheme, when it is applied to real astronomical data, where the true parameter values and specific noise realisation are unknown. \newline

@@

Figure \ref{fig:corner_plot_2} confirms two main points (i) the results from the nine noise realisations overlap with the single realisation in  Figure \ref{fig:corner_plot_1}; and (ii) the dispersion among the peaks of the one-dimensional posteriors is appreciable, with variations of $\sim10 \%$ (coefficient of variation) across the seven parameters and nine realizations. Indeed, considering the one-dimensional marginalized posteriors, we find that the injected value is contained within the 90\% credible interval in 58 out of the 63 (i.e. $92 \%$) possible combinations of the seven parameters and nine realizations. Figure \ref{fig:corner_plot_2}, like Figure \ref{fig:corner_plot_1}, displays tentative signs of bias, where the one-dimensional posteriors are not symmetric about the injected value. For example, the maximum a posteriori probability estimates of $\iota$ appear left-skewed, consistently underestimating the injected value in all nine realisations. A similar but messier trend is seen in $\delta$. However, the width of the posteriors is comparable to the putative bias, so it is difficult to draw strong conclusions. Bias is discussed in detail in Section \ref{sec:bias}. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours. \newline

We now consider the complete set of $10^3$ noise realisations, going beyond than the subset of nine realisations in the preceding discussion. We ask the question: how ``similar'' are the $10^3$ marginalized, one-dimensional posteriors computed for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$? There is no unique way to answer this question. In this paper we appeal to the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory, which defines an intuitive notion of similarity between probability distributions. The WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}; a short review is presented in Appendix \ref{sec:wasserstein}. The WD measures the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. In this paper we use the first WD moment, $W_1(\mu,\nu)$, which is defined and interpreted in Appendix \ref{sec:wasserstein}. Intuitively $W_1$ bounds the difference in the expectation value of a parameter selected from ${\boldsymbol{\theta}}_{\rm gw}$" with respect to the PDFs $\mu$ and $\nu$. Taking a concrete example, suppose that we infer two one-dimensional posterior distributions $\mu(\iota)$ and $\nu(\iota)$ for $\iota$, for two different realisations of the noise, and calculate $W_1(\mu, \nu) =0.5$ rad. Then we can infer $| \langle \iota \rangle_\mu - \langle \iota \rangle_\nu | \leq 0.5 \, {\rm rad}$. \newline

Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$. As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown. Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.14 \times 10^{-4}$ \% for $h_0$ to 2.69\% for $\psi$. These values serve as confirmation that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations. Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.

\subsection{Detectability versus $h_0$} \label{sec:detection}
We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}. In Equatin \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model, i.e. the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}. The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable). To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself. \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, simply adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline

We see in Figure \ref{fig:bayes} an approximate quadratic relationship $\beta \propto h_0^2$ exists. The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 4 \times 10^{-15}$. Of course, the minimal detectable strain is particular to the system in Table \ref{tab:parameters_and_priors}. It is influenced in general by e.g. $T_{\rm obs}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$, as discussed in Section \ref{sec:parameter_space}. Adjusting $\sigma_{\rm m}$ linearly scales the effective signal-to-noise ratio $h_0 / \sigma_{\rm m}$, such that an approximate quadratic relationship $\beta \propto h_0^2 / \sigma_{\rm m}^2$ also exists. \newline

The odds ratio $\beta$ drops off for $h_0 \lesssim 4 \times10^{-15}$. This happens because the two competing models becoming increasingly indistinguishable once the measurement noise dominates the GW signal. Moreover, the points in Figure \ref{fig:bayes} become sparser for $h_0 \lesssim 4 \times 10^{-15}$. This happens because we find $\beta <0$ for the missing, intermediate points. This is a noise artefact of the nested sampler.When the sampler converges sub-optimally, the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails, and $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds. There is nothing special about the particular missing points; if one recreates the $\beta(h_0)$ curve by rerunning the nested sampler with another random seed, a different set of points are missing. Increasing the number of live points $n_{\rm live}$ used by the nested sampler improves the performance at small values of $h_0$, because the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(n_{\rm live}^{-1/2})$ (see Section \ref{sec:nested_sampling}). In this section we set $n_{\rm live} =2000$. 
\begin{figure}
	\includegraphics[width=\columnwidth]{images/CanonicalBayesPlot2000}
	\caption{Bayes factor (odds ratio) $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW amplitudes, $h_0$, for the representative example in Table \ref{tab:parameters_and_priors}. The horizontal grey dashed line labels an arbitrary detection threshold, $\beta = 10$. The minimum detectable strain for $\beta < 10$, is $ 4 \times 10^{-15}$. Missing points for $\beta\leq 10$ occur when noise subverts the hierarchical relationship between $\mathcal{M}_0$ and $\mathcal{M}_1$ (see Section \ref{sec:detection}).}
	\label{fig:bayes}
\end{figure}

@@

\section{Identifiability and bias} \label{sec:bias_and_identifiability}
There are two underlying issues in the preceding tests that we now explore in more detail. The first relates to parameter identifiability, i.e. can we uniquely identify the parameters of the model given the data and the model structure? This question is explored in Section \ref{sec:identif}. The second relates to a bias in the parameter estimates that results from dropping the pulsar terms from the Kalman filter measurement equation as described in Section \ref{sec:parameter_estim}. This is the focus of Section \ref{sec:bias}. In order to elucidate these questions without confusion from the measurement noise, we now work in the high SNR regime and set $h_0 = 10^{-12}$ (c.f. Figure \ref{fig:bayes}) for our representative example system (Table \ref{tab:parameters_and_priors}).

@@

\section{Wasserstein distance}\label{sec:wasserstein}
The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$. It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$. For this reason it is sometimes known as the "Earth mover's distance". The $p$-th order Wasserstein distance between two distributions is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
\end{eqnarray}
where $p \in [1,\infty)$, $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e. $\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.

@@

The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages. Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another. It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g. radians then we can interpret this as the cost to transform between the two distributions is 1 radian. Further interpretability is provided by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. That is, if $W_1 =1$ then the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian. Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g. Gaussianity. Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces. Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality. It also respects the geometry of the underlying metric space over which the distributions are defined. \newline

%%%%%%%%%%%%%%%

\begin{figure*}
	\setkeys{Gin}{width=\linewidth}

	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_0}
		\caption{$\Omega$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_1}
		\caption{$\Phi_0$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_2}
		\caption{$\psi$} \label{figWD_psi}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics[width=\textwidth]{images/WD_3}
		\caption{$\iota$}
	\end{subfigure}
	\medskip
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_4}
		\caption{$\delta$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_5}
		\caption{$\alpha$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/WD_6}
		\caption{$h$}
	\end{subfigure}
	\caption{The first moment of the Wasserstein distance, $W_1$, calculated between each pair of one-dimensional posteriors for our representative system across $10^3$ realisations, for each parameter of $\boldsymbol{\theta}_{\rm gw}$  $W_1$ provides an upper bound on the difference in expected values between any two probability distributions. $W_1$ is generally small across all parameters and all posteriors, but some posteriors converge poorly leading to large values of $W_1$. The $W_1$ values inferred for $\psi$ and $\alpha$ are strongly correlated with correlation coefficient = 0.96. An example noise realisation with particularly large pairwise $W_1$ values is labelled with an asterisk in (c) and is discussed in the text.} \label{fig:pairwise_wasserstein}
\end{figure*}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface_alpha_psi}
	\caption{Log-likelihood, $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\alpha-\psi$ parameter space for a single noise realisation.  The red point labels the location of the maxima, and the magenta point labels the median values of the one-dimensional posteriors inferred for a particular noise realisation of the data. The likelihood has been normalised with respect to the absolute value of the maxima. A multi-modal structure is clearly visible and for this noise realisation the nested sampler has settled in the local optima. This can be solved generally by increasing the number of live points $n_{\rm live}$.}
	\label{fig:likelihood_surface_alpha_psi}
\end{figure}
